---
title: "Global Options"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Global Options}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```
The **procs** packages contains some global options that may be helpful 
in some scenarios.  

### Turn On/Off Interactive Report 

By default, several **procs** functions print an interactive report to the 
RStudio viewer.  For instance, `proc_freq()`, `proc_ttest()`, and `proc_reg()`
all print tables and even plots interactively. These functions all have options to
restrict printing, usually called "noprint".  

However, sometimes it is more
convenient to restrict printing for all **procs** functions, rather than
turn them off one by one. In this scenario, you can turn the interactive display
off globally in the following manner:
```{r eval=FALSE, echo=TRUE}
library(procs)

# Turn printing off
options("procs.print" = FALSE)

```

The above option will restrict printing for any **procs** package function.

To restore the interactive reports, either set the "procs.print" value to TRUE,
or remove the option entirely by setting it to NULL:
```{r eval=FALSE, echo=TRUE}
# Turn printing on 
options("procs.print" = TRUE)

# Delete printing option
options("procs.print" = NULL)

```

### Control Sorting Behavior

The `proc_sort()` function sorts a data frame.  By default, the function
sorts the "by" parameter variables in ascending order, and NA values are 
sorted to the bottom.  The "na.sort" parameter contains options on how 
NA values are sorted.  The parameter takes four possible values: 
NULL, "first", "last", and "sas".  The "first" and "last" settings will
sort NA values first or last, respectively.  The "sas" setting will
sort NA values first for ascending sorts and last for descending sorts.

You can control the sort behavior globally with the "procs.na.sort" option.
Like this:
```{r eval=FALSE, echo=TRUE}
# Sort NAs first
options("procs.na.sort" = "first")

# Sort NAs last
options("procs.print" = "last")

# Sort to match SAS
options("procs.print" = "sas")

```

The NULL setting on `proc_sort()` means to defer to the global setting.
Otherwise, the local setting will override the global setting.  

To remove the global setting, set the option to NULL:
```{r eval=FALSE, echo=TRUE}
# Delete global setting
options("procs.na.sort" = NULL)

```

### Log Printing

If you are using the **procs** package in combination with the **logr** package
for logging, **procs** functions can print to the log automatically. You may
allow or prevent log printing with the option "logr.output":

```{r eval=FALSE, echo=TRUE}
# Prevent automatic log printing 
options("logr.output" = FALSE)

# Allow automatic log printing 
options("logr.output" = TRUE)

# Delete logging option
options("logr.output" = NULL)

```
Note that to enable automatic logging, you must also set the "logr.autolog" 
option to TRUE.  The "logr.output" option only controls the automatic logging of
functions within the [SASSY](https://r-sassy.org/) system of packages, and only
when "logr.autolog" = TRUE.
