---
title: "Frequently Asked Questions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(rmarkdown.html_vignette.check_title = FALSE)

```

Below are some frequently asked questions about the **libr** package. Click
on the links below to navigate to the full question and answer content.

## Index{#top}

* [How do I create a libname in R?](#libname)
* [Which data formats does the libname function support?](#formats)



## Content

### How do I create a libname in R? {#libname}

**Q:** I have a directory full of datasets.  I need to use several of them
in my analysis.  In SAS®, I would create a libname so I could access all of them.
Is there a way to do something similar in R?

**A:** With the **libr** package, you can create a libname in R very 
much the same way you create a libname in SAS®.  
```{r eval=FALSE, echo=TRUE}
libname(mylib, "c:/mypath/mydata", "csv")

```
The above statement will create a libname "mylib" from the directory
specified on the second parameter.  The libname will use the CSV engine.
If there are any CSV files in the directory, they will be all loaded 
into the library.  To work directly with the datasets, you can then do:
```{r eval=FALSE, echo=TRUE}
lib_load(mylib)
```
This statement will load the datasets into your workspace, where you can begin
using them.  For instance, you could get summary statistics for a variable like
this:
```{r eval=FALSE, echo=TRUE}
summary(mylib.dat1$var1)
```

If you made any changes to the data, and want to keep those changes, remember
to save them with:
```{r eval=FALSE, echo=TRUE}
lib_write(mylib)
```
When you are done, unload the datasets with:
```{r eval=FALSE, echo=TRUE}
lib_unload(mylib)
```

[top](#top) 

******

### Which data formats does the libname function support? {#formats}

**Q:** I can see from the examples that the **libr** package supports
CSV and SAS dataset file formats.  What other data formats does the package
support?


**A:** The package supports the following data formats: csv, sas7bdat, 
rds, Rdata, xls, xlsx, xpt, and dbf.  The `libname()` help page has a full list,
and a short discussion of some details on each format.

[top](#top) 

******


<!-- ### Question 2? {#q2} -->

<!-- **Q:** Question here. -->

<!-- **A:** Answer here. -->

<!-- [top](#top) -->

<!-- ****** -->

<!-- ### Question 1? {#q1} -->

<!-- **Q:** Question here. -->


<!-- **A:** Answer here. -->

<!-- [top](#top)  -->

<!-- ****** -->

