---
title: "procs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{procs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **procs** package contains functions that replicate procedures from
SAS® software. The intention of the package is to ease transition to R
adoption by providing SAS® programmers a familiar conceptual framework and 
functions that produce nearly identical output. Along the way, the 
functions in the **procs** package also provide much nicer output 
than existing R statistical functions.

### Key Functions

The package includes the following major functions:

* `proc_freq()`: A function to simulate the FREQ procedure.
* `proc_means()`: A function to simulate the MEANS or SUMMARY procedure.
* `proc_transpose()`: A function to pivot data similar in syntax to the TRANSPOSE procedure.


### How to Use


The following examples illustrate the basic functionality of each of the 
above three functions:

#### The `proc_freq()` Function

The `proc_freq()` function generates frequency statistics in a manner
similar to SAS® PROC FREQ.  

```{r eval=FALSE, echo=TRUE} 
library(procs)


```

#### The `proc_means()` Function

The `proc_means()` function attempts to provide similar functionality to 
the SAS® PROC MEANS procedure.

```{r eval=FALSE, echo=TRUE} 
library(procs)


```

#### The `proc_transpose()` Function

While the main purpose of the **procs** package is to replicate SAS® statistical
procedures, the output from these procedures is often manipulated to
produce a more desirable result.  While R and Tidyverse do contain sufficient
data manipulation functions, the TRANSPOSE procedure from SAS® provides a 
somewhat different conceptual model that SAS® programmers like.  The aim
of the `proc_transpose()` function is to offer a similar model in R.

```{r eval=FALSE, echo=TRUE} 
library(procs)


```

### Next Steps

For next steps, please review the examples provided in the vignette articles. 
Those articles include:

* [The Frequency Function](procs-freq.html)
* [The Means Function](procs-means.html)
* [The Transpose Function](procs-transpose.html)
* [Package Validation](procs-validation.html)
* [Complete Example](procs-example1.html)




